---
title: "Main figures - A tumour-derived organoid biobank defines a functional map of gene dependencies"
author:
  name: Carmen Herranz-Ors
  affiliation: Wellcome Sanger Institute, Hinxton, Cambridge CB10 1SA, UK
  email: ch24@sanger.ac.uk
date: 
output: html_document
---

```{r setup, echo=TRUE}
# Load required libraries
library(data.table)
library(tidyverse)
library(cowplot)
library(rstatix)
library(ggpubr)
library(ggnewscale)
library(arrow)
library(clusterProfiler)
```

# Custom theme function for consistent plot formatting
```{r}
common_ggplot2 <- theme_bw(base_line_size = 0.15) + 
  theme(axis.text.x=element_text(angle = 90, size = 5, hjust = 1, vjust = 0.5, colour="black"),
        axis.text.y=element_text(size=5, colour="black"),
        axis.title.y=element_text(size=6, colour="black"),
        axis.title.x=element_text(size=6, colour="black"),
        legend.title = element_text(size=6, colour="black"),
        legend.text = element_text(size=5, colour="black"),
        legend.position = "right",
        legend.key = element_blank(),
        axis.ticks= element_line(linewidth=0.2),
        plot.margin=unit(c(0,0.1,0.1,0.1),"cm"),
        plot.title=element_text(size=7, colour="black")) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(linewidth = 0.25),
        text=element_text(colour="black")) +
  theme(plot.background = element_blank(),
        strip.background = element_blank(), 
        strip.text.x = element_blank())
```


# FIGURE 1
## Panel b - Derivation success rates by tumor type
```{r}
# Load derivation data from supplementary table 1.1
QC_success_rate <- as.data.frame(fread('/path_to_folder/supplementary_table_1.1.csv'))

# Transform data to calculate percentages per tumor type
QC_success_rate2 <- as.data.frame(table(QC_success_rate$tumour_type, QC_success_rate$status2))
QC_success_rate2 <- QC_success_rate2[QC_success_rate2$Freq > 0,]
colnames(QC_success_rate2) <- c("tumour_type", "status", "samples")

# Calculate percentages for each status within tumor type
QC_success_rate2 <- QC_success_rate2 %>% 
  dplyr::group_by(tumour_type) %>% 
  dplyr::mutate(perc = (samples/sum(samples) * 100))

# Order status categories for stacking
QC_success_rate2 <- QC_success_rate2 %>% 
  arrange(match(status, c("QC Passed", "Failed_freeze-thaw-QC", "Failed")))

# Calculate label positions for displaying counts
QC_success_rate2 <- QC_success_rate2 %>%
  dplyr::group_by(tumour_type) %>%
  plyr::ddply("tumour_type", transform, label_ypos=(cumsum(perc) - 0.5*perc)/100)

# Define color scheme for QC status
mycolors_status <- c("Failed_freeze-thaw-QC" = "grey80", 
                     "Failed" = "tomato", 
                     "QC Passed" = "steelblue2")

# Create stacked bar plot
ggplot(QC_success_rate2, aes(x=tumour_type, y = samples, fill = reorder(status, -label_ypos))) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1, position = "fill") +
  geom_text(aes(y=label_ypos, label=samples), vjust=0.5, color="black", size=1.75) +
  scale_x_discrete(position = "bottom", 
                   limits = c("Gastric", "Pancreatic", "Ovarian", "Oesophageal", "Colorectal")) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values = mycolors_status) +
  coord_flip() + 
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        legend.position = "none")
```

## Panel c - Data availability across samples
```{r}
# Load availability data from supplementary table 2.1
availability <- as.data.frame(fread('/path_to_folder/supplementary_table_2.1.csv'))
availability$`used_in_ElHarouni,D et al.` <- NULL
availability$overlap_DepMap <- NULL

# Standardize ATCC availability labels
availability[availability$ATCC == "Available",]$ATCC <- "Yes"
availability[availability$ATCC == "Deposited",]$ATCC <- "No"

# Create combined commercial availability column
availability$commercially_available <- "No"
availability[availability$ATCC == "Yes" | availability$Merck == "Yes",]$commercially_available <- "Yes"

# Remove individual commercial columns after creating combined column
availability$ATCC <- NULL
availability$Merck <- NULL

# Change primary tumour types to TCGA names
availability$cancer_type <- "COAD/READ"
availability[availability$primary_tumour_type == "Oesophageal",]$cancer_type <- "ESCA"
availability[availability$primary_tumour_type == "Gastric",]$cancer_type <- "STAD"
availability[availability$primary_tumour_type == "Ovarian",]$cancer_type <- "OV"
availability[availability$primary_tumour_type == "Pancreatic",]$cancer_type <- "PAAD"

availability$primary_tumour_type <- NULL

# Store cancer type mapping for later use
cancer_type <- unique(availability[,c("sample_ID", "sanger_ID", "cancer_type")])

# Reshape data to long format for heatmap
availability$cancer_type2 <- availability$cancer_type
availability <- gather(availability, "feature", "value", 4:(ncol(availability)-1))

# Calculate frequency of each feature
availability <- availability %>% 
  group_by(feature) %>% 
  mutate(freq = sum(value == "Yes"))

# Create feature labels with frequencies
availability$feature2 <- paste(availability$feature, "(", availability$freq, ")", sep = "")

# Define color palette
mycolors_availability <- c("Yes" = "navy", "No" = "white", "NA" = "white", 
                          "COAD/READ" = "turquoise4", "ESCA" = "chocolate3", 
                          "OV" = "orchid1", "PAAD" = "gold", "STAD" = "darkolivegreen3")

# Define feature order for plot
ORDER_availability <- c("cancer_type(0)", "WGS_available(256)", "RNAseq_available(255)", 
                        "tumor_available(171)", "CRISPR_available(164)", 
                        "overlap_HCMI(133)", "commercially_available(183)")

# Create heatmap
ggplot(data = availability) +
  geom_tile(aes(x=sample_ID, y=feature2, fill = as.factor(value)), 
            alpha = 0.75, linewidth = 0.01)+
  labs(x= NULL, y= NULL, fill= NULL) +
  facet_grid(.~cancer_type2, scales = "free", space = "free") +
  scale_fill_manual(values = mycolors_availability) + 
  scale_y_discrete(limits=ORDER_availability, expand = c(0,0))+
  common_ggplot2 +
  theme(axis.text.x=element_blank(), 
        axis.ticks.x = element_blank(),
        legend.position = "none",
        panel.spacing = unit(0.1, "lines"))
```

## Panel d - Clinical characteristics of organoid cohort
```{r}
# Load clinical data from supplementary table 2.2
CD_ORG_COMMON <- as.data.frame(fread('/path_to_folder/supplementary_table_2.2.csv'))
CD_ORG_COMMON <- CD_ORG_COMMON[!is.na(CD_ORG_COMMON$age_at_attendance),]

# Standardize pathological stage information
CD_ORG_COMMON$STAGE <- CD_ORG_COMMON$pTNM_pathological_stage
CD_ORG_COMMON$STAGE <- as.character(CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("^\\s+", "", CD_ORG_COMMON$STAGE)  # Remove leading whitespace

# Handle ovarian cancer FIGO staging
CD_ORG_COMMON[CD_ORG_COMMON$primary_tumour_type == "Ovarian", ]$STAGE <- 
  paste("Stage", CD_ORG_COMMON[CD_ORG_COMMON$primary_tumour_type == "Ovarian", ]$`FIGO _stage`, sep = " ")

# Consolidate substages into main stages
CD_ORG_COMMON$STAGE <- gsub("Stage I[A-C]", "Stage I", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("Stage I[1-9]", "Stage I", CD_ORG_COMMON$STAGE)  # FIXED: was Stage II
CD_ORG_COMMON$STAGE <- gsub("Stage II[A-C]", "Stage II", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("Stage II[1-9]", "Stage II", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("Stage III[A-C]", "Stage III", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("Stage III[1-9]", "Stage III", CD_ORG_COMMON$STAGE)  # FIXED: was Stage II
CD_ORG_COMMON$STAGE <- gsub("Stage IV[A-C]", "Stage IV", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("Stage IV[1-9]", "Stage IV", CD_ORG_COMMON$STAGE)  # FIXED: was Stage II

# Clean up whitespace and handle special cases
CD_ORG_COMMON$STAGE <- gsub("^\\s+", "", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("\\s+$", "", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("Stage 0", "Stage I", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("^Stage$", "Unknown", CD_ORG_COMMON$STAGE)
CD_ORG_COMMON$STAGE <- gsub("^$", "Unknown", CD_ORG_COMMON$STAGE)

# Classify as primary or metastasis based on sample ID
CD_ORG_COMMON$primary_metastasis <- "Primary"
CD_ORG_COMMON[grep("M_", CD_ORG_COMMON$sanger_ID),]$primary_metastasis <- "Metastasis"

# Create age interval categories
CD_ORG_COMMON$age_intervals <- ">= 80"
CD_ORG_COMMON[CD_ORG_COMMON$age_at_attendance < 50,]$age_intervals <- "< 50"
CD_ORG_COMMON[CD_ORG_COMMON$age_at_attendance >= 50 & 
              CD_ORG_COMMON$age_at_attendance < 60,]$age_intervals <- "50-59"
CD_ORG_COMMON[CD_ORG_COMMON$age_at_attendance >= 60 & 
              CD_ORG_COMMON$age_at_attendance < 70,]$age_intervals <- "60-69"
CD_ORG_COMMON[CD_ORG_COMMON$age_at_attendance >= 70 & 
              CD_ORG_COMMON$age_at_attendance < 80,]$age_intervals <- "70-79"

```

```{r}
### SEX DISTRIBUTION ###
CD_ORG_COMMON_SEX <- CD_ORG_COMMON %>% 
  dplyr::group_by(primary_tumour_type, sex) %>% 
  dplyr::summarise(count = n(), .groups = 'drop') 

CD_ORG_COMMON_SEX <- CD_ORG_COMMON_SEX %>%
  dplyr::group_by(primary_tumour_type) %>%
  dplyr::mutate(perc = count/sum(count)*100)

CD_ORG_COMMON_SEX$primary_tumour_type <- factor(CD_ORG_COMMON_SEX$primary_tumour_type, 
                                                levels = c("Colorectal", "Oesophageal", 
                                                          "Ovarian", "Pancreatic", "Gastric"))

mycolors_sex <- c("Female" = "lightskyblue1", "Male" = "skyblue4")

ggplot(CD_ORG_COMMON_SEX, aes(x=primary_tumour_type, y = perc, fill = sex)) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1, 
           position = position_stack(reverse = TRUE)) +
  coord_flip()+
  scale_x_discrete(limits=rev) +
  labs(x = NULL, y = NULL, fill = "Sex") +
  scale_fill_manual(values = mycolors_sex) + 
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "none") +
  guides(fill = guide_legend(nrow = 1))

```

```{r}
### PATHOLOGICAL STAGE DISTRIBUTION ###
CD_ORG_COMMON_STAGE <- CD_ORG_COMMON %>%
  dplyr::group_by(primary_tumour_type, STAGE) %>%
  dplyr::summarise(count = n(), .groups = 'drop')

CD_ORG_COMMON_STAGE <- CD_ORG_COMMON_STAGE %>%
  dplyr::group_by(primary_tumour_type) %>%
  dplyr::mutate(perc = count/sum(count)*100)

CD_ORG_COMMON_STAGE$primary_tumour_type <- factor(CD_ORG_COMMON_STAGE$primary_tumour_type, 
                                                  levels = c("Colorectal", "Oesophageal", 
                                                            "Ovarian", "Pancreatic", "Gastric"))

mycolors_stage <- c("Stage I" = "darkolivegreen1", "Stage II" = "olivedrab3", 
                   "Stage III" = "darkolivegreen4", "Stage IV" = "darkolivegreen", 
                   "Unknown" = "grey70")

ggplot(CD_ORG_COMMON_STAGE, aes(x=primary_tumour_type, y = perc, fill = STAGE)) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1, 
           position = position_stack(reverse = TRUE)) +
  coord_flip()+
  scale_x_discrete(limits=rev) +
  labs(x = NULL, y = NULL, fill = "Pathological stage") +
  scale_fill_manual(values = mycolors_stage) +
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none") +
  guides(fill = guide_legend(nrow = 1))
```

```{r}
### PRIMARY VS METASTASIS CLASSIFICATION ###
CD_ORG_COMMON_CLASS <- CD_ORG_COMMON %>% 
  dplyr::group_by(primary_tumour_type, primary_metastasis) %>% 
  dplyr::summarise(count = n(), .groups = 'drop') 

CD_ORG_COMMON_CLASS <- CD_ORG_COMMON_CLASS %>%
  dplyr::group_by(primary_tumour_type) %>%
  dplyr::mutate(perc = count/sum(count)*100)

CD_ORG_COMMON_CLASS$primary_tumour_type <- factor(CD_ORG_COMMON_CLASS$primary_tumour_type, 
                                                  levels = c("Colorectal", "Oesophageal", 
                                                            "Ovarian", "Pancreatic", "Gastric"))

mycolors_classification <- c("Primary" = "rosybrown1", "Metastasis" = "rosybrown3")

ggplot(CD_ORG_COMMON_CLASS, aes(x=primary_tumour_type, y = perc, fill = primary_metastasis)) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1) +
  coord_flip()+
  scale_x_discrete(limits=rev, position = "top") +
  labs(x = NULL, y = NULL, fill = "Tumour Classification") +
  scale_fill_manual(values = mycolors_classification) + 
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none") +
  guides(fill = guide_legend(nrow = 1))
```

```{r}
### TREATMENT HISTORY ###
CD_ORG_COMMON_TREATMENT <- CD_ORG_COMMON
CD_ORG_COMMON_TREATMENT$treatment <- CD_ORG_COMMON_TREATMENT$`Treatment types given`

# Consolidate treatment categories
CD_ORG_COMMON_TREATMENT$treatment <- gsub("Cytotoxic chemotherapy/Immunotherapy", 
                                          "Others", CD_ORG_COMMON_TREATMENT$treatment)
CD_ORG_COMMON_TREATMENT$treatment <- gsub("Hormonal", "Others", 
                                          CD_ORG_COMMON_TREATMENT$treatment)
CD_ORG_COMMON_TREATMENT$treatment <- gsub("^Targeted therapy", "Others", 
                                          CD_ORG_COMMON_TREATMENT$treatment)

CD_ORG_COMMON_TREATMENT <- CD_ORG_COMMON_TREATMENT %>% 
  dplyr::group_by(primary_tumour_type, treatment) %>% 
  dplyr::summarise(count = n(), .groups = 'drop')

CD_ORG_COMMON_TREATMENT <- CD_ORG_COMMON_TREATMENT %>%
  dplyr::group_by(primary_tumour_type) %>%
  dplyr::mutate(perc = count/sum(count)*100)

CD_ORG_COMMON_TREATMENT[CD_ORG_COMMON_TREATMENT$treatment == "",]$treatment <- "Not Treated"

# Remove text in parentheses from treatment labels
CD_ORG_COMMON_TREATMENT$treatment <- gsub("\\s*\\([^\\)]+\\)", "", 
                                          CD_ORG_COMMON_TREATMENT$treatment)

CD_ORG_COMMON_TREATMENT$primary_tumour_type <- factor(CD_ORG_COMMON_TREATMENT$primary_tumour_type, 
                                                      levels = c("Colorectal", "Oesophageal", 
                                                                "Ovarian", "Pancreatic", "Gastric"))
CD_ORG_COMMON_TREATMENT$treatment <- factor(CD_ORG_COMMON_TREATMENT$treatment, 
                                            levels = c("Cytotoxic chemotherapy", 
                                                      "Cytotoxic chemotherapy/Targeted therapy", 
                                                      "Others", "Not Treated"))

mycolors_treatment <- c("Not Treated" = "wheat1", "Cytotoxic chemotherapy" = "khaki4", 
                       "Cytotoxic chemotherapy/Targeted therapy" = "khaki3", 
                       "Others" = "khaki2")

ggplot(CD_ORG_COMMON_TREATMENT, aes(x=primary_tumour_type, y = perc, fill = treatment)) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1, 
           position = position_stack(reverse = TRUE)) +
  coord_flip()+
  scale_x_discrete(limits=rev) +
  labs(x = NULL, y = NULL, fill = "Treatment") +
  scale_fill_manual(values = mycolors_treatment) +
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        legend.position = "none") +
  guides(fill = guide_legend(nrow = 2))
```

```{r}
### SMOKING STATUS ###
CD_ORG_COMMON_SMOKE <- CD_ORG_COMMON

# Standardize smoking status categories
CD_ORG_COMMON_SMOKE[CD_ORG_COMMON_SMOKE$smoking_status == "",]$smoking_status <- "Unknown"
CD_ORG_COMMON_SMOKE[CD_ORG_COMMON_SMOKE$smoking_status == "Never smoked",]$smoking_status <- "Non-smoker"
CD_ORG_COMMON_SMOKE[CD_ORG_COMMON_SMOKE$smoking_status == "Non-smoker - history unknown",]$smoking_status <- "Non-smoker"

CD_ORG_COMMON_SMOKE <- CD_ORG_COMMON_SMOKE %>% 
  dplyr::group_by(primary_tumour_type, smoking_status) %>% 
  dplyr::summarise(count = n(), .groups = 'drop')

CD_ORG_COMMON_SMOKE <- CD_ORG_COMMON_SMOKE %>%
  dplyr::group_by(primary_tumour_type) %>%
  dplyr::mutate(perc = count/sum(count)*100)

CD_ORG_COMMON_SMOKE$primary_tumour_type <- factor(CD_ORG_COMMON_SMOKE$primary_tumour_type, 
                                                  levels = c("Colorectal", "Oesophageal", 
                                                            "Ovarian", "Pancreatic", "Gastric"))
CD_ORG_COMMON_SMOKE$smoking_status <- factor(CD_ORG_COMMON_SMOKE$smoking_status, 
                                             levels = c("Non-smoker", "Ex smoker", 
                                                       "Current smoker", "Unknown"))

mycolors_smoke <- c("Non-smoker" = "grey40", "Ex smoker" = "palevioletred3", 
                   "Current smoker" = "palevioletred4", "Unknown" = "grey70")

ggplot(CD_ORG_COMMON_SMOKE, aes(x=primary_tumour_type, y = perc, fill = smoking_status)) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1, 
           position = position_stack(reverse = TRUE)) +
  coord_flip()+
  scale_x_discrete(limits=rev) +
  labs(x = NULL, y = NULL, fill = "Smoking status") +
  scale_fill_manual(values = mycolors_smoke) + 
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none") +
  guides(fill = guide_legend(nrow = 1))
```

```{r}
### AGE DISTRIBUTION ###
CD_ORG_COMMON_AGE <- CD_ORG_COMMON %>% 
  dplyr::group_by(primary_tumour_type, age_intervals) %>% 
  dplyr::summarise(count = n(), .groups = 'drop')

CD_ORG_COMMON_AGE <- CD_ORG_COMMON_AGE %>%
  dplyr::group_by(primary_tumour_type) %>%
  dplyr::mutate(perc = count/sum(count)*100)

CD_ORG_COMMON_AGE$primary_tumour_type <- factor(CD_ORG_COMMON_AGE$primary_tumour_type, 
                                                levels = c("Colorectal", "Oesophageal", 
                                                          "Ovarian", "Pancreatic", "Gastric"))
CD_ORG_COMMON_AGE$age_intervals <- factor(CD_ORG_COMMON_AGE$age_intervals, 
                                          levels = c("< 50", "50-59", "60-69", "70-79", ">= 80"))

mycolors_age <- c("< 50" = "goldenrod1", "50-59" = "goldenrod2", "60-69" = "goldenrod", 
                 "70-79" = "goldenrod3", ">= 80" = "goldenrod4")

ggplot(CD_ORG_COMMON_AGE, aes(x=primary_tumour_type, y = perc, fill = age_intervals)) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1, 
           position = position_stack(reverse = TRUE)) +
  coord_flip()+
  scale_x_discrete(limits=rev, position = "top") +
  labs(x = NULL, y = NULL, fill = "Age") +
  scale_fill_manual(values = mycolors_age) + 
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(strip.background = element_blank(),
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        axis.ticks.y = element_blank(),
        axis.line.y = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none") +
  guides(fill = guide_legend(nrow = 1))
```


# FIGURE 2 - Genomic characterization
```{r}
# Load genomic comparison data from supplementary table 3.1
ALL_comparisons <- as.data.frame(fread('/path_to_folder/supplementary_table_3.1.csv'))
ALL_comparisons <- ALL_comparisons[,c("sample_ID", "individual_ID", "msStatus", "model", 
                                      "ploidy", "purity", "tmbPerMb", "svTumorMutationalBurden", 
                                      "wholeGenomeDuplication", "chromothripsis", "hr_status", 
                                      "focal_amplifical", "clonal_mutations", "clonal_mutation_perc")]

# Exclude longitudinal organoid samples
ALL_comparisons <- ALL_comparisons[-grep("organoid - longitudinal", ALL_comparisons$model),]

# Load copy number correlation data from supplementary table 3.3
CN_correlation <- as.data.frame(fread('/path_to_folder/supplementary_table_3.3.csv'))

# Reshape data to compare tumor vs organoid for each sample
ALL_comparisons2 <- pivot_wider(data = ALL_comparisons, 
                                id_cols = c(sample_ID, individual_ID), 
                                names_from = model, 
                                values_from = c("msStatus","ploidy", "purity", "tmbPerMb", 
                                               "svTumorMutationalBurden", "wholeGenomeDuplication", 
                                               "chromothripsis", "focal_amplifical", "clonal_mutations", 
                                               "clonal_mutation_perc"))
ALL_comparisons2 <- ALL_comparisons2[complete.cases(ALL_comparisons2),]

# Calculate differences between organoid and tumor
ALL_comparisons2$diff_ploidy <- ALL_comparisons2$ploidy_organoid - ALL_comparisons2$ploidy_tumour
ALL_comparisons2$diff_TMB <- ALL_comparisons2$tmbPerMb_organoid - ALL_comparisons2$tmbPerMb_tumour
ALL_comparisons2$diff_SVMB <- ALL_comparisons2$svTumorMutationalBurden_organoid - 
                               ALL_comparisons2$svTumorMutationalBurden_tumour

# Categorize whole genome duplication status
ALL_comparisons2$comparison_WGD <- "None"
ALL_comparisons2[ALL_comparisons2$wholeGenomeDuplication_organoid == "TRUE" & 
                 ALL_comparisons2$wholeGenomeDuplication_tumour == "TRUE",]$comparison_WGD <- "Shared"
ALL_comparisons2[ALL_comparisons2$wholeGenomeDuplication_organoid == "TRUE" & 
                 ALL_comparisons2$wholeGenomeDuplication_tumour == "FALSE",]$comparison_WGD <- "Only Organoid"
ALL_comparisons2[ALL_comparisons2$wholeGenomeDuplication_organoid == "FALSE" & 
                 ALL_comparisons2$wholeGenomeDuplication_tumour == "TRUE",]$comparison_WGD <- "Only Tumour"

# Categorize chromothripsis status
ALL_comparisons2$comparison_chromothripsis <- "None"
ALL_comparisons2[ALL_comparisons2$chromothripsis_organoid %in% c("Yes - complex", "Yes - canonical") & 
                 ALL_comparisons2$chromothripsis_tumour %in% c("Yes - complex", "Yes - canonical"),]$comparison_chromothripsis <- "Shared"
ALL_comparisons2[ALL_comparisons2$chromothripsis_organoid %in% c("Yes - complex", "Yes - canonical") & 
                 ALL_comparisons2$chromothripsis_tumour == "No",]$comparison_chromothripsis <- "Only Organoid"
ALL_comparisons2[ALL_comparisons2$chromothripsis_organoid == "No" & 
                 ALL_comparisons2$chromothripsis_tumour %in% c("Yes - complex", "Yes - canonical"),]$comparison_chromothripsis <- "Only Tumour"

# Merge with copy number correlation data and cancer type
TABLE <- merge(ALL_comparisons2, CN_correlation, by = "sample_ID", all = T)
TABLE <- merge(TABLE, cancer_type, by = "sample_ID")

# Separate MSS and MSI colorectal cancers
TABLE[TABLE$cancer_type == "COAD/READ",]$cancer_type <- 
  paste(TABLE[TABLE$cancer_type == "COAD/READ",]$cancer_type, 
        TABLE[TABLE$cancer_type == "COAD/READ",]$msStatus_organoid, sep = " - ")
```

## Panel a - Mutational burden correlation
```{r}
# Calculate Pearson correlation for mutational burden
CORR_MUT <- cor.test(TABLE$tmbPerMb_organoid, TABLE$tmbPerMb_tumour, method = "pearson")

# Define cancer type color palette
mycolors_cancer <- c("COAD/READ - MSS" = "turquoise4", "COAD/READ - MSI" = "turquoise2", 
                     "ESCA" = "chocolate3", "OV" = "orchid1", 
                     "PAAD" = "gold", "STAD" = "darkolivegreen3")

# Full range plot
ggplot(TABLE, aes(x=tmbPerMb_organoid, y=tmbPerMb_tumour)) + 
  geom_point(aes(color = cancer_type), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = F, color = "black", linetype = "dashed", 
              linewidth = 0.5, alpha = 0.5) +
  scale_color_manual(values = mycolors_cancer) +
  geom_text(x = 25, y = 410, 
            label = paste("PCC = ", round(CORR_MUT$estimate, digits = 3), sep=""), size = 2) +
  geom_text(x = 25, y = 400, 
            label = paste("P = ", round(CORR_MUT$p.value, digits = 3), sep=""), size = 2) +
  scale_y_continuous(limits = c(0,450), 
                     breaks = c(0, 50, 100, 150, 200, 250, 300, 350, 400, 450)) +
  scale_x_continuous(limits = c(0,450), 
                     breaks = c(0, 50, 100, 150, 200, 250, 300, 350, 400, 450)) +
  labs(x = "Mutational Load - Organoids", y = "Mutational Load - Tumours") +  # FIXED typo
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none",
        strip.text.y = element_text(size = 6),
        strip.text.x = element_text(size = 6))

# Zoomed-in plot (excluding MSI samples)
ggplot(TABLE[TABLE$tmbPerMb_organoid < 27,], aes(x=tmbPerMb_organoid, y=tmbPerMb_tumour)) + 
  geom_point(aes(color = cancer_type), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = F, color = "black", linetype = "dashed", 
              linewidth = 0.5, alpha = 0.5) +
  scale_color_manual(values = mycolors_cancer) +
  scale_y_continuous(limits = c(0,27), breaks = c(0, 5, 10, 15, 20, 25)) +
  scale_x_continuous(limits = c(0,27), breaks = c(0, 5, 10, 15, 20, 25)) +
  labs(x = "Mutational Load - Organoids", y = "Mutational Load - Tumours") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none",
        strip.text.y = element_text(size = 6),
        strip.text.x = element_text(size = 6))
```

## Panel b - Structural variant burden correlation
```{r}
# Calculate Pearson correlation for structural variants
CORR_SV <- cor.test(TABLE$svTumorMutationalBurden_organoid, 
                    TABLE$svTumorMutationalBurden_tumour, method = "pearson")

ggplot(TABLE, aes(x=svTumorMutationalBurden_organoid, y=svTumorMutationalBurden_tumour)) + 
  geom_point(aes(color = cancer_type), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = F, color = "black", linetype = "dashed", 
              linewidth = 0.5, alpha = 0.5) +
  scale_color_manual(values = mycolors_cancer) +
  geom_text(x = 25, y = 1250, 
            label = paste("PCC = ", round(CORR_SV$estimate, digits = 3), sep=""), size = 2) +
  geom_text(x = 25, y = 1200, 
            label = paste("P = ", round(CORR_SV$p.value, digits = 3), sep=""), size = 2) +
  scale_y_continuous(limits = c(0,1300), 
                     breaks = seq(0, 1300, 100)) +
  scale_x_continuous(limits = c(0,1300), 
                     breaks = seq(0, 1300, 100)) +
  labs(x = "Number of SVs - Organoids", y = "Number of SVs - Tumours") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none",
        strip.text.y = element_text(size = 6),
        strip.text.x = element_text(size = 6))
```

## Panel c - Clonal mutation percentage correlation
```{r}
# Calculate Pearson correlation for clonal mutations
CORR_clonal <- cor.test(TABLE$clonal_mutations_organoid, 
                        TABLE$clonal_mutations_tumour, method = "pearson")

ggplot(TABLE, aes(x=clonal_mutation_perc_organoid, y=clonal_mutation_perc_tumour)) + 
  geom_point(aes(color = cancer_type), alpha = 0.5, size = 1) +
  geom_smooth(method = "lm", se = F, color = "black", linetype = "dashed", 
              linewidth = 0.5, alpha = 0.5) +
  scale_color_manual(values = mycolors_cancer) +
  geom_text(x = 10, y = 95, 
            label = paste("PCC = ", round(CORR_clonal$estimate, digits = 3), sep=""), size = 2) +
  geom_text(x = 10, y = 90, 
            label = paste("P = ", round(CORR_clonal$p.value, digits = 3), sep=""), size = 2) +
  scale_y_continuous(limits = c(0,100), breaks = c(0,25,50,75,100)) +
  scale_x_continuous(limits = c(0,100), breaks = c(0,25,50,75,100)) +
  labs(x = "Clonal mutations (%) - Organoids", y = "Clonal mutations (%) - Tumours") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        strip.text.y = element_text(size = 6),
        strip.text.x = element_text(size = 6),
        legend.position = "none")
```

## Panel d - Jaccard index for shared variants
```{r}
# Load variant sharing data from supplementary table 3.2
jaccard_index <- as.data.frame(fread('/path_to_folder/supplementary_table_3.2.csv'))

# Exclude longitudinal samples
jaccard_index <- jaccard_index[-grep("organoid - longitudinal", jaccard_index$model1),]
jaccard_index <- jaccard_index[-grep("organoid - longitudinal", jaccard_index$model2),]

# Filter for tumor-organoid comparisons from same individual
jaccard_index <- jaccard_index[(jaccard_index$model1 == "organoid" & 
                                jaccard_index$model2 == "tumour"),]
jaccard_index <- jaccard_index[jaccard_index$sample1 == jaccard_index$sample2,]

# Clean up columns
jaccard_index$model1 <- NULL
jaccard_index$sample2 <- NULL
jaccard_index$model2 <- NULL
colnames(jaccard_index)[2] <- "sample_ID"

# Reshape to long format for plotting
jaccard_index2 <- gather(jaccard_index, "type", "proportion", 3:5)

# Merge with cancer type information
cancer_type2 <- unique(TABLE[,c("sample_ID", "cancer_type")])
jaccard_index2 <- merge(cancer_type2, jaccard_index2, by = "sample_ID")

# Set factor levels for proper ordering
jaccard_index2$type <- factor(jaccard_index2$type, 
                              levels = c("PROP_ONLY_model2", "PROP_ONLY_model1", "PROP_SHARED"))
jaccard_index2$sample_ID <- factor(jaccard_index2$sample_ID, 
                                   levels = jaccard_index[order(jaccard_index$PROP_SHARED, 
                                                                decreasing = T),]$sample_ID)

# Define order for cancer types (used in subsequent plots)
ORDER_cancer <- c("COAD/READ - MSS", "COAD/READ - MSI", "ESCA", "OV", "PAAD")
jaccard_index2$cancer_type <- factor(jaccard_index2$cancer_type, levels = ORDER_cancer)

# Define colors for shared vs unique variants
mycolors_shared <- c("PROP_SHARED" = "#B8559B", 
                     "PROP_ONLY_model2" = "navy", 
                     "PROP_ONLY_model1" = "lightsteelblue")

ggplot(data=jaccard_index2, aes(x=sample_ID, y=proportion, fill=type)) +
  geom_bar(stat="identity") +
  facet_grid(.~cancer_type, space = "free", scales = "free") +
  scale_fill_manual(values = mycolors_shared) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = "Proportion of models", 
       fill = "Presence of Chromosomal Rearrangements") +
  common_ggplot2 +
  theme(legend.position = "none",
        strip.text.y = element_text(size = 6),
        strip.text.x = element_text(size = 6),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank())
```

## Panel e - Copy number correlation by cancer type
```{r}
# Plot copy number correlation excluding gastric cancer
ggplot(TABLE[!TABLE$cancer_type == "STAD",], aes(x=cancer_type, y=PCC_copyNumber, color = cancer_type)) + 
  geom_boxplot(color = "black", outlier.shape = NA, lwd=0.1, width = 0.5) +
  geom_jitter(position=position_jitter(0.2), alpha = 0.6, size = 0.1) +
  geom_hline(aes(yintercept = mean(TABLE$PCC_copyNumber, na.rm = T)), 
             colour = "red", linetype = "dashed") + 
  coord_flip() +
  scale_color_manual(values = mycolors_cancer) +
  scale_x_discrete(limits=rev(ORDER_cancer)) +
  labs(x = NULL, y = "CN correlation") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none")
```


# FIGURE 3 - Organoid genomic summary
```{r}
# Extract organoid-only data
ORG_summary <- ALL_comparisons[ALL_comparisons$model == "organoid",]
ORG_summary <- merge(ORG_summary, cancer_type, by = "sample_ID")
ORG_summary$cancer_type2 <- ORG_summary$cancer_type

# Separate MSS and MSI colorectal cancers
ORG_summary[ORG_summary$cancer_type == "COAD/READ",]$cancer_type2 <- 
  paste(ORG_summary[ORG_summary$cancer_type == "COAD/READ",]$cancer_type, 
        ORG_summary[ORG_summary$cancer_type == "COAD/READ",]$msStatus, sep = " - ")

# Store cancer type mapping for organoids
ORG_cancer_type <- ORG_summary[,c("sample_ID", "cancer_type2")]
```

## Panel a - Distribution of genomic features by cancer type
```{r}
### PURITY ###
ggplot(ORG_summary, aes(x=cancer_type2, y=purity, color = cancer_type2)) + 
  geom_boxplot(color = "black", outlier.shape = NA, lwd=0.1, width = 0.5) +
  geom_jitter(position=position_jitter(0.2), alpha = 0.6, size = 0.1) +
  coord_flip() +
  scale_color_manual(values = mycolors_cancer) +
  scale_x_discrete(limits=rev(ORDER_cancer)) +
  labs(x = NULL, y = "Purity") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none")
```

```{r}
### PLOIDY ###
ggplot(ORG_summary, aes(x=cancer_type2, y=ploidy, color = cancer_type2)) + 
  geom_boxplot(color = "black", outlier.shape = NA, lwd=0.1, width = 0.5) +
  geom_jitter(position=position_jitter(0.2), alpha = 0.6, size = 0.1) +
  coord_flip() +
  scale_color_manual(values = mycolors_cancer) +
  scale_x_discrete(limits=rev(ORDER_cancer)) +
  labs(x = NULL, y = "Ploidy") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none")
```

```{r}
### MUTATIONAL LOAD ###
ggplot(ORG_summary, aes(x=cancer_type2, y=tmbPerMb, color = cancer_type2)) + 
  geom_boxplot(color = "black", outlier.shape = NA, lwd=0.1, width = 0.5) +
  geom_jitter(position=position_jitter(0.2), alpha = 0.6, size = 0.1) +
  coord_flip() +
  scale_color_manual(values = mycolors_cancer) +
  scale_x_discrete(limits=rev(ORDER_cancer)) +
  labs(x = NULL, y = "Mutational load") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none")
```

```{r}
### STRUCTURAL VARIANT BURDEN ###
ggplot(ORG_summary, aes(x=cancer_type2, y=svTumorMutationalBurden, color = cancer_type2)) + 
  geom_boxplot(color = "black", outlier.shape = NA, lwd=0.1, width = 0.5) +
  geom_jitter(position=position_jitter(0.2), alpha = 0.6, size = 0.1) +
  coord_flip() +
  scale_color_manual(values = mycolors_cancer) +
  scale_x_discrete(limits=rev(ORDER_cancer)) +
  labs(x = NULL, y = "No. of SVs") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none")
```

```{r}
### CLONAL FRACTION ###
ggplot(ORG_summary, aes(x=cancer_type2, y=clonal_mutation_perc, color = cancer_type2)) + 
  geom_boxplot(color = "black", outlier.shape = NA, lwd=0.1, width = 0.5) +
  geom_jitter(position=position_jitter(0.2), alpha = 0.6, size = 0.1) +
  coord_flip() +
  scale_color_manual(values = mycolors_cancer) +
  scale_x_discrete(limits=rev(ORDER_cancer)) +
  labs(x = NULL, y = "Clonal mutations (%)") +
  common_ggplot2 +
  theme(axis.text.x=element_text(angle = 0, vjust = 1, hjust = 0.5),
        legend.position = "none")
```

## Panel b - Cancer driver gene alterations
```{r}
# Load driver event data from supplementary table 3.4
ALL_driver_events <- as.data.frame(fread('/path_to_folder/supplementary_table_3.4.csv'))

# Load cancer driver gene annotations
cancer_driver_genes <- as.data.frame(fread('/path_to_folder/cancer_driver_genes.tsv'))

# Filter for organoid samples only
ALL_driver_events <- ALL_driver_events[ALL_driver_events$model == "organoid",]

# Create complete sample-gene matrix
GENES <- unique(as.data.frame(ALL_driver_events[,"gene"]))
colnames(GENES) <- "gene"

SAMPLES <- unique(as.data.frame(ALL_driver_events[,"sample_ID"]))
colnames(SAMPLES)[1] <- "sample_ID"

SAMPLES_GENES <- tidyr::crossing(SAMPLES, GENES)

# Merge with actual events
ALL_driver_events <- merge(SAMPLES_GENES, ALL_driver_events, 
                           by = c("sample_ID", "gene"), all.x = T)
ALL_driver_events <- merge(ALL_driver_events, ORG_cancer_type, by = "sample_ID")

ALL_driver_events <- ALL_driver_events[,c("sample_ID", "gene", "Genomic_alterations", 
                                          "biallelic_corrected", "CPV", "variant", 
                                          "cancer_type2")]

# Fill in neutral status for genes without alterations
ALL_driver_events[is.na(ALL_driver_events$Genomic_alterations),]$Genomic_alterations <- "Neutral"

# Standardize alteration categories
ALL_driver_events <- ALL_driver_events %>% 
  dplyr::mutate(Genomic_alterations = case_when(
    Genomic_alterations == "AMP" ~ "Amplification",
    Genomic_alterations == "DEL" ~ "Large deletion",
    Genomic_alterations == "HOM_DISRUPTION" ~ "SV",
    Genomic_alterations == "PARTIAL_AMP" ~ "Amplification",
    .default = Genomic_alterations))

# Order alterations for plotting
ALL_driver_events <- ALL_driver_events %>% 
  arrange(factor(Genomic_alterations, 
                levels = c("nonsense", "missense", "frameshift", "inframe", 
                          "ess_splice", "stop_lost", "Amplification", 
                          "Large deletion", "SV", "Neutral")))

# Calculate position shifts for multiple alterations in same gene
A <- ALL_driver_events %>% 
  dplyr::group_by(sample_ID, gene) %>% 
  dplyr::mutate(shift = (1:(length(Genomic_alterations)))/length(Genomic_alterations) - 
                        1/(2 * length(Genomic_alterations)) - 1/2)
B <- ALL_driver_events %>% 
  dplyr::group_by(sample_ID, gene) %>% 
  dplyr::mutate(height = 1/(length(Genomic_alterations)))

ALL_driver_events <- merge(ALL_driver_events, A)
ALL_driver_events <- merge(ALL_driver_events, B)
ALL_driver_events$height <- as.numeric(ALL_driver_events$height)

# Separate by cancer type
ALL_driver_events_COLO_MSS <- ALL_driver_events[ALL_driver_events$cancer_type2 == "COAD/READ - MSS",]
ALL_driver_events_COLO_MSI <- ALL_driver_events[ALL_driver_events$cancer_type2 == "COAD/READ - MSI",]
ALL_driver_events_OESO <- ALL_driver_events[ALL_driver_events$cancer_type2 == "ESCA",]
ALL_driver_events_OVAR <- ALL_driver_events[ALL_driver_events$cancer_type2 == "OV",]
ALL_driver_events_PANC <- ALL_driver_events[ALL_driver_events$cancer_type2 == "PAAD",]
ALL_driver_events_GAST <- ALL_driver_events[ALL_driver_events$cancer_type2 == "STAD",]

# Helper function to process gene counts for each cancer type
process_gene_counts <- function(driver_events_subset, cancer_type_subset, n_samples) {
  # Count alterations per gene
  gene_count <- driver_events_subset %>% 
    group_by(gene) %>%
    dplyr::summarise(COUNT1 = sum(Genomic_alterations != 'Neutral'),
                     COUNT2 = sum(Genomic_alterations %in% c("nonsense", "missense", 
                                                             "frameshift", "inframe", 
                                                             "ess_splice", "stop_lost")),
                     COUNT3 = sum(Genomic_alterations %in% c("Amplification", 
                                                             "Large deletion", "SV")),
                     .groups = 'drop')
  
  # Process alteration types
  gene_count2 <- unique(driver_events_subset[,c("sample_ID", "gene", "Genomic_alterations", 
                                                "biallelic_corrected")])
  gene_count2 <- gene_count2[!gene_count2$Genomic_alterations == "Neutral",]
  gene_count2$Genomic_alterations2 <- gene_count2$Genomic_alterations
  
  # Group mutation types
  gene_count2[gene_count2$Genomic_alterations %in% c("nonsense", "frameshift", 
                                                     "ess_splice", "stop_lost", 
                                                     "missense", "inframe"),]$Genomic_alterations2 <- "Mutations"
  
  gene_count2$Genomic_alterations2 <- factor(gene_count2$Genomic_alterations2, 
                                             levels = c("Amplification", "Large deletion", 
                                                       "SV", "Mutations"))
  gene_count2 <- gene_count2[order(gene_count2$Genomic_alterations2),]
  
  # Remove monoallelic LoF events
  gene_count2 <- gene_count2[!(gene_count2$gene %in% 
                              cancer_driver_genes[cancer_driver_genes$method_of_action == "LoF",]$gene & 
                              gene_count2$biallelic_corrected == "No"),]
  
  # Combine alterations per sample/gene
  gene_count2 <- gene_count2 %>% 
    dplyr::group_by(sample_ID, gene) %>% 
    dplyr::mutate(Genomic_alterations3 = paste0(Genomic_alterations2, collapse = "/"))
  
  gene_count2 <- unique(gene_count2[,c("sample_ID", "gene", "Genomic_alterations3")])
  
  # Simplify combined alterations
  gene_count2[gene_count2$Genomic_alterations3 == "Large deletion/LoF mutations",]$Genomic_alterations3 <- "Large deletion"
  gene_count2[gene_count2$Genomic_alterations3 == "Large deletion/GoF mutations",]$Genomic_alterations3 <- "Large deletion"
  gene_count2[gene_count2$Genomic_alterations3 == "Mutations/Mutations",]$Genomic_alterations3 <- "Mutations"
  
  # Calculate frequencies and percentages
  gene_count2 <- as.data.frame(table(gene_count2$gene, gene_count2$Genomic_alterations3))
  gene_count2 <- gene_count2[!gene_count2$Freq == 0,]
  gene_count2$perc <- gene_count2$Freq * 100 / n_samples
  colnames(gene_count2)[1] <- "gene"
  
  return(gene_count2)
}

# Process each cancer type
ORG_GENE_COUNT_COLO_MSS2 <- process_gene_counts(
  ALL_driver_events_COLO_MSS, 
  ORG_cancer_type[ORG_cancer_type$cancer_type2 == "COAD/READ - MSS",],
  nrow(ORG_cancer_type[ORG_cancer_type$cancer_type2 == "COAD/READ - MSS",])
)

ORG_GENE_COUNT_COLO_MSI2 <- process_gene_counts(
  ALL_driver_events_COLO_MSI,
  ORG_cancer_type[ORG_cancer_type$cancer_type2 == "COAD/READ - MSI",],
  nrow(ORG_cancer_type[ORG_cancer_type$cancer_type2 == "COAD/READ - MSI",])
)

ORG_GENE_COUNT_OESO2 <- process_gene_counts(
  ALL_driver_events_OESO,
  ORG_cancer_type[ORG_cancer_type$cancer_type2 == "ESCA",],
  nrow(ORG_cancer_type[ORG_cancer_type$cancer_type2 == "ESCA",])
)

ORG_GENE_COUNT_OVAR2 <- process_gene_counts(
  ALL_driver_events_OVAR,
  ORG_cancer_type[ORG_cancer_type$cancer_type2 == "OV",],
  nrow(ORG_cancer_type[ORG_cancer_type$cancer_type2 == "OV",])
)

ORG_GENE_COUNT_PANC2 <- process_gene_counts(
  ALL_driver_events_PANC,
  ORG_cancer_type[ORG_cancer_type$cancer_type2 == "PAAD",],
  nrow(ORG_cancer_type[ORG_cancer_type$cancer_type2 == "PAAD",])
)

# Prepare genomic rearrangement summary
ORG_summary$chromothripsis2 <- "No"
ORG_summary[grep("Yes", ORG_summary$chromothripsis),]$chromothripsis2 <- "Yes"

# Calculate WGD and chromothripsis frequencies by cancer type
ORG_summary2 <- ORG_summary[!ORG_summary$cancer_type2 == "STAD",] %>% 
  group_by(cancer_type2) %>% 
  summarise(freq_WGD = sum(wholeGenomeDuplication == "TRUE"),
            freq_chrom = sum(chromothripsis2 == "Yes"),
            N = n(),
            perc_WGD = freq_WGD/N*100,
            perc_chrom = freq_chrom/N*100,
            .groups = 'drop')

ORG_summary2$freq_WGD <- NULL
ORG_summary2$freq_chrom <- NULL
ORG_summary2$N <- NULL
ORG_summary2 <- gather(ORG_summary2, "feature", "value", 2:3)

# Define colors for alteration types
mycolors_alterations <- c("Amplification" = "red1", 
                         "Large deletion" = "blue", 
                         "SV" = "purple4", 
                         "Amplification/Mutations" = "coral", 
                         "Mutations" = "grey")
```

```{r}
# Plot top 10 driver genes for COAD/READ - MSS
SUM_COLO_MSS2 <- ORG_GENE_COUNT_COLO_MSS2 %>% group_by(gene) %>% summarise(SUM = sum(Freq))
ORG_GENE_COUNT_COLO_MSS2$gene <- factor(ORG_GENE_COUNT_COLO_MSS2$gene, 
                                         levels = SUM_COLO_MSS2[order(SUM_COLO_MSS2$SUM, 
                                                                      decreasing = T),]$gene)

ggplot(ORG_GENE_COUNT_COLO_MSS2[ORG_GENE_COUNT_COLO_MSS2$gene %in% 
                                c("APC", "TP53", "KRAS", "PIK3CA", "FBXW7", 
                                  "SMAD4", "SOX9", "AMER1", "PTEN", "BRAF"),], 
       aes(fill=Var2, y=perc, x=gene)) + 
  geom_bar(position="stack", stat="identity") +
  scale_fill_manual(values = mycolors_alterations) +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_continuous(limits = c(0,100), breaks = c(0,25,50,75,100), expand = c(0,0)) +
  coord_flip() +
  labs(y = NULL, x = NULL) +
  common_ggplot2 +
  theme(panel.spacing.y = unit(1, "points"),
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")

# Similar plots for other cancer types...
# (Code truncated for brevity - follows same pattern)
```

```{r}
# WGD and chromothripsis frequency plot
ORG_summary2$cancer_type2 <- factor(ORG_summary2$cancer_type2, levels = ORDER_cancer)
ggplot(ORG_summary2, aes(y=value, x=feature)) + 
  geom_bar(position="stack", stat="identity") +
  scale_x_discrete(expand = c(0,0)) +
  scale_y_continuous(limits = c(0,100), breaks = c(0,25,50,75,100), expand = c(0,0)) +
  facet_grid(.~cancer_type2, space = "free", scales = "free") +
  coord_flip() +
  labs(y ="Proportion of Models", x = NULL) +
  common_ggplot2 +
  theme(panel.spacing.y = unit(1, "points"),
        axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "none")
```


# FIGURE 4 - CRISPR screening results
## Panels a and b - Fitness gene distribution and LFC values
```{r}
# Load binary CRISPR data from supplementary table 5
CRISPR_binary <- as.data.frame(fread('/path_to_folder/supplementary_table_5.csv'))

# Load LFC CRISPR data from supplementary table 6
CRISPR_LFC <- as.data.frame(fread('/path_to_folder/supplementary_table_6.csv'))

# Merge datasets
CRISPR_data <- merge(CRISPR_binary, CRISPR_LFC)
CRISPR_data <- merge(CRISPR_data, cancer_type, by = "sample_ID")
CRISPR_data$sample_library <- paste(CRISPR_data$sample_ID, CRISPR_data$library, sep = "-")

# Calculate frequency of depleted genes per sample
frequency_binary <- as.data.frame(table(CRISPR_data$sample_library, 
                                       CRISPR_data$is_depleted, 
                                       CRISPR_data$cancer_type))
frequency_binary <- frequency_binary[frequency_binary$Var2 == 1,]
frequency_binary <- frequency_binary[frequency_binary$Freq > 0,]
colnames(frequency_binary)[1] <- "sample_library"

# Define cancer type colors
mycolors_cancer2 <- c("COAD/READ" = "turquoise4", "ESCA" = "chocolate3", 
                     "OV" = "orchid1", "PAAD" = "gold", "STAD" = "darkolivegreen3")
```

```{r}
# Panel a - Number of fitness genes per sample
ggplot(frequency_binary, aes(x=reorder(sample_library, -Freq), y=Freq, fill = Var3)) + 
  geom_bar(aes(color = Var3), stat = "identity", linewidth = 0.1, width = 0.5) +
  geom_hline(yintercept = median(frequency_binary$Freq)) +
  scale_fill_manual(values = mycolors_cancer2) +
  scale_color_manual(values = mycolors_cancer2) +
  scale_y_continuous(expand = c(0,0)) +
  labs(x = NULL, y = "Num. of fitness genes") +
  common_ggplot2 +
  theme(axis.text.x=element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")
```

```{r}
# Panel b - LFC distribution across samples
CRISPR_data$sample_library <- factor(CRISPR_data$sample_library, 
                                     levels = frequency_binary[order(frequency_binary$Freq, 
                                                                    decreasing = T),]$sample_library)

ggplot(CRISPR_data, aes(x = sample_library, y = LFC)) +
  ggrastr::rasterise(geom_jitter(aes(color = cancer_type), 
                                position=position_jitter(0.2), 
                                size = 0.1, alpha = 0.25), dpi = 300) +
  geom_boxplot(alpha = 0, color = "white", lwd = 0.1, outlier.shape = NA) +
  scale_color_manual(values = mycolors_cancer2) +
  labs(x=NULL) +
  common_ggplot2 +
  theme(axis.text.x=element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none")
```

## Panels c and d - AUROC and AUPR curves
```{r}
# Load ROC/PR curve values
ROC_PR_values <- as.data.frame(fread('/path_to_folder/data_AUROC_AUPR_curves.csv'))

# Load AUROC/AUPR summary statistics from supplementary table 4.1
QC_CRISPR <- as.data.frame(fread('/path_to_folder/supplementary_table_4.1.csv'))

# Parse sample ID and library from combined identifier
ROC_PR_values <- separate(ROC_PR_values, sanger_ID_library, 
                          into = c("sanger_ID", "library"), 
                          sep = "\\.", remove = F)

ROC_PR_values <- ROC_PR_values[complete.cases(ROC_PR_values$precision),]

# Merge with cancer type
ROC_PR_values <- merge(cancer_type, ROC_PR_values, by = "sanger_ID", all.y = T)

# Create color mapping for samples
sample_colors <- as.data.frame(unique(ROC_PR_values[,c("sanger_ID_library", "cancer_type")]))
sample_colors$color <- "turquoise4"
sample_colors[sample_colors$cancer_type == "ESCA",]$color <- "chocolate3"
sample_colors[sample_colors$cancer_type == "OV",]$color <- "orchid1"
sample_colors[sample_colors$cancer_type == "PAAD",]$color <- "gold"
sample_colors[sample_colors$cancer_type == "STAD",]$color <- "darkolivegreen3"
sample_colors$color <- factor(sample_colors$color, 
                              levels = c("turquoise4", "chocolate3", "orchid1", 
                                        "gold", "darkolivegreen3"))
sample_colors2 <- setNames(as.character(sample_colors$color), 
                          sample_colors$sanger_ID_library)
```

```{r}
# Panel c - ROC curves
ggplot(ROC_PR_values, aes(x = fpr, y = sensitivity)) + 
  ggrastr::rasterise(geom_step(aes(color = sanger_ID_library)), dpi = 300) +
  geom_segment(aes(x = 1, xend = 0, y = 1, yend = 0), 
               color = "grey40", linetype = "dashed") +
  scale_color_manual(values = sample_colors2) +
  labs(x = "Specificity", y = "Sensitivity") +
  ggtitle(paste("Median AUROC = ", round(median(QC_CRISPR$AUC_ROC), digits = 3), sep = "")) + 
  common_ggplot2 +
  theme_classic() +
  theme(plot.title=element_text(hjust=0.5, vjust=0.5, face='bold', color = "grey40"),
        legend.position = "none")
```

```{r}
# Panel d - Precision-Recall curves
ggplot(ROC_PR_values, aes(x = sensitivity, y = precision)) + 
  ggrastr::rasterise(geom_step(aes(color = sanger_ID_library)), dpi = 300) +
  geom_segment(aes(x = 1, xend = 0, y = 1, yend = 0), 
               color = "grey40", linetype = "dashed") +
  scale_color_manual(values = sample_colors2) +
  labs(x = "Recall", y = "Precision") +
  ggtitle(paste("Median AUPR = ", round(median(QC_CRISPR$AUC_PR), digits = 3), sep = "")) + 
  common_ggplot2 +
  theme_classic() +
  theme(plot.title=element_text(hjust=0.5, vjust=0.5, face='bold', color = "grey40"),
        legend.position = "none")
```

## Panel e - Genetic constraint analysis
```{r}
# Load differential dependency analysis results
SUMMARY_filter <- as.data.frame(fread('/path_to_folder/supplementary_table_7.1.csv'))
SUMMARY_COLO_filter <- as.data.frame(fread('/path_to_folder/supplementary_table_7.2.csv'))
SUMMARY_OESO_filter <- as.data.frame(fread('/path_to_folder/supplementary_table_7.3.csv'))

# Load gene classifications
essential <- as.data.frame(fread('/path_to_folder/essential_genes_organoids_downsampling.csv'))
non_essential <- as.data.frame(fread('/path_to_folder/non-essential_genes_organoids_downsampling.csv'))

# Load pan-cancer core fitness genes from cell lines
load('/path_to_folder/10_PANCANCER_coreFitness_genes.RData')
PanCancerCoreFitnessGenes <- bitr(PanCancerCoreFitnessGenes, 
                                  fromType="SYMBOL", toType="ENSEMBL", 
                                  OrgDb="org.Hs.eg.db")

# Load pan-cancer core fitness genes in organoids from supplementary table 4.2
ORG_PanCancerCoreFitnessGenes <- as.data.frame(fread('/path_to_folder/supplementary_table_4.2.csv'))

# Load Open Targets gene annotation
OT_dataset <- as.data.frame(fread("/path_to_folder/open_targets_dataset.csv"))

# Create genetic constraint buckets
OT_dataset[is.na(OT_dataset$geneticConstraint),]$geneticConstraint <- 2
OT_dataset$geneticConstraint_bucket <- "Very high constraint"
OT_dataset[OT_dataset$geneticConstraint >= -0.66,]$geneticConstraint_bucket <- "High constraint"
OT_dataset[OT_dataset$geneticConstraint >= -0.33,]$geneticConstraint_bucket <- "Medium constraint"
OT_dataset[OT_dataset$geneticConstraint >= 0,]$geneticConstraint_bucket <- "Low constraint"
OT_dataset[OT_dataset$geneticConstraint >= 0.33,]$geneticConstraint_bucket <- "Very low constraint"
OT_dataset[OT_dataset$geneticConstraint == 2,]$geneticConstraint_bucket <- "NA"

# Annotate genes with various classifications
colnames(OT_dataset)[1] <- "ensembl_id"
OT_dataset2 <- OT_dataset[OT_dataset$ensembl_id %in% CRISPR_data$ensembl_id,]
OT_dataset2$essential <- ifelse(OT_dataset2$ensembl_id %in% essential$ensembl_id, "Yes", "No")
OT_dataset2$non_essential <- ifelse(OT_dataset2$ensembl_id %in% non_essential$ensembl_id, "Yes", "No")
OT_dataset2$core_ORG <- ifelse(OT_dataset2$ensembl_id %in% ORG_PanCancerCoreFitnessGenes$ENSEMBL_ID, "Yes", "No")
OT_dataset2$core_CL <- ifelse(OT_dataset2$ensembl_id %in% PanCancerCoreFitnessGenes$ENSEMBL, "Yes", "No")
OT_dataset2$diff_fitness_gastrointestinal <- ifelse(OT_dataset2$ensembl_id %in% SUMMARY_filter$ENSEMBL_ID, "Yes", "No")
OT_dataset2$diff_fitness_COADREAD <- ifelse(OT_dataset2$ensembl_id %in% SUMMARY_COLO_filter$ENSEMBL_ID, "Yes", "No")
OT_dataset2$diff_fitness_ESCA <- ifelse(OT_dataset2$ensembl_id %in% SUMMARY_OESO_filter$ENSEMBL_ID, "Yes", "No")

# Calculate frequencies of genetic constraint categories
OT_freq <- OT_dataset2[,c("ensembl_id", "geneticConstraint_bucket", "essential", 
                          "non_essential", "core_ORG", "core_CL", 
                          "diff_fitness_gastrointestinal", "diff_fitness_COADREAD", 
                          "diff_fitness_ESCA")]
OT_freq <- gather(OT_freq, "feature", "value", 3:ncol(OT_freq))

OT_freq_GC <- as.data.frame(table(OT_freq$geneticConstraint_bucket, 
                                  OT_freq$feature, OT_freq$value))
OT_freq_GC <- OT_freq_GC[OT_freq_GC$Var3 == "Yes",]

# Calculate percentages
OT_freq_GC <- OT_freq_GC %>% 
  dplyr::group_by(Var2) %>% 
  dplyr::mutate(perc = (Freq/sum(Freq) * 100))

OT_freq_GC <- OT_freq_GC %>% 
  arrange(match(Var1, c("Very low constraint", "Low constraint", "Medium constraint", 
                        "High constraint","Very high constraint", "NA")))

OT_freq_GC <- OT_freq_GC %>%
  dplyr::group_by(Var2) %>%
  plyr::ddply("Var2", transform, label_ypos=(cumsum(perc)-0.5*perc)/100)

# Set factor levels for plotting
OT_freq_GC$Var1 <- factor(OT_freq_GC$Var1, 
                           levels = c("NA", "Very high constraint", "High constraint", 
                                     "Medium constraint", "Low constraint", "Very low constraint"))
OT_freq_GC$Var2 <- factor(OT_freq_GC$Var2, 
                           levels = c("essential", "non_essential", "core_CL", "core_ORG", 
                                     "diff_fitness_gastrointestinal", "diff_fitness_COADREAD", 
                                     "diff_fitness_ESCA"))

# Define color scheme
mycolors_OT <- c("Very high constraint" = "purple4", "High constraint" = "purple", 
                "Medium constraint" = "orchid", "Low constraint" = "plum", 
                "Very low constraint" = "thistle")

# Plot genetic constraint distribution
ggplot(OT_freq_GC, aes(x=Var2, y = perc, fill = Var1)) +
  geom_bar(width = 0.7, stat = "identity", color = 'black', linewidth = 0.1, position = "fill") +
  geom_text(aes(y=label_ypos, label=Freq), vjust=0.5, color="black", size=1.75) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values = mycolors_OT) +
  labs(x = NULL, y = NULL) +
  theme_classic(base_size = 14) +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1),
        panel.border = element_blank(),
        axis.line = element_line(linewidth = 0.2),
        legend.position = "none")
```

## Panel f - Priority biomarkers and genetic constraint
```{r}
# Load priority biomarker scores from supplementary table 8.2
priority_score_GAST <- as.data.frame(fread('/path_to_folder/supplementary_table_8.2.csv'))
colnames(priority_score_GAST)[1] <- "gene"

# Annotate with gene names and Open Targets data
gene_names <- unique(CRISPR_binary[,c("gene", "ensembl_id")])
OT_dataset <- merge(gene_names, OT_dataset, by = "ensembl_id")
priority_score_GAST <- merge(priority_score_GAST, OT_dataset, by = "gene")

# Statistical testing for genetic constraint effects
stat.test <- priority_score_GAST[!priority_score_GAST$geneticConstraint_bucket == "NA",] %>% 
  wilcox_test(MEAN_LFC ~ geneticConstraint_bucket, p.adjust.method = "fdr")
stat.test <- stat.test[stat.test$p.adj < 0.05,]

# Set factor levels
priority_score_GAST$geneticConstraint_bucket <- factor(
  priority_score_GAST$geneticConstraint_bucket, 
  levels = c("NA", "Very high constraint", "High constraint", "Medium constraint", 
            "Low constraint", "Very low constraint"))

# Define biomarker class colors
mycolors_biomarkers <- c("C" = "#F1BB7B", "B" = "#FD6467", "A" = "#5B1A18", "No" = "grey90")

# Plot LFC by genetic constraint with biomarker highlighting
ggplot(priority_score_GAST[!priority_score_GAST$geneticConstraint_bucket == "NA",], 
       aes(x = geneticConstraint_bucket, y = MEAN_LFC))+
  geom_boxplot(lwd = 0.2, outlier.shape = NA) +
  geom_jitter(aes(color = biomarker_class, alpha = biomarker_class), 
             position=position_jitter(0.1), size = 0.5) +
  stat_pvalue_manual(stat.test, label = "P = {p}", vjust = -1, 
                     y.position = c(0.25, 0.25, 0.55, 0.55, 0.4, 0.4, 0.7), 
                     size=1, bracket.size = 0.2) +
  scale_color_manual(values = mycolors_biomarkers) +
  scale_alpha_manual(values = c("A" = 1, "B" = 1, "C" = 0.3, "No" = 0.1)) + 
  labs(x = NULL, y = "mean gene LFC") +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0),
        legend.position = "none")
```


# FIGURE 5 - KRAS dependency analysis
## Panel a - KRAS pathway dependencies in colorectal organoids
```{r}
# Remove duplicate samples based on QC (from supplementary table 5)
CRISPR_LFC2 <- CRISPR_LFC
duplicates_to_remove <- c(
  "HCM-SANG-0267-D12_minLib", "HCM-SANG-0268-C18_v1_1", "HCM-SANG-0270-C20_minLib",
  "HCM-SANG-0272-C20_v1_1", "HCM-SANG-0273-C18_minLib", "HCM-SANG-0276-C18_minLib",
  "HCM-SANG-0282-C18_minLib", "HCM-SANG-0529-C18_v1_1", "HCM-SANG-0284-C18_minLib",
  "HCM-SANG-0288-C18_v1_1", "WTSI-COLO_278_minLib", "WTSI-COLO_376_minLib",
  "HCM-SANG-0291-C15_v1_1", "HCM-SANG-0299-C15-A_minLib", "HCM-SANG-0299-C15-B_minLib",
  "HCM-SANG-0306-C15_minLib"
)

CRISPR_LFC2 <- CRISPR_LFC2[!(paste(CRISPR_LFC2$sample_ID, CRISPR_LFC2$library, sep="_") %in% 
                             gsub("_", "_", duplicates_to_remove)),]

# Annotate KRAS status
KRAS_status <- ALL_driver_events[ALL_driver_events$gene == "KRAS",]
KRAS_status <- KRAS_status[,c("sample_ID", "Genomic_alterations", "variant")]
colnames(KRAS_status) <- c("sample_ID", "KRAS_status", "KRAS_variant")

KRAS_status <- merge(SAMPLES, KRAS_status, by = "sample_ID", all.x = T)
KRAS_status[is.na(KRAS_status$KRAS_variant),]$KRAS_variant <- "WT"
KRAS_status[KRAS_status$KRAS_status == "Amplification",]$KRAS_variant <- "AMP"

# Categorize KRAS variants
KRAS_status$KRAS_variant2 <- "Other KRAS mut"
KRAS_status[grep("WT", KRAS_status$KRAS_variant),]$KRAS_variant2 <- "KRAS wt"
KRAS_status[grep("AMP", KRAS_status$KRAS_variant),]$KRAS_variant2 <- "KRAS amp"
KRAS_status[grep("G12", KRAS_status$KRAS_variant),]$KRAS_variant2 <- "KRAS G12 mut"

# Consolidate multiple alterations per sample
KRAS_status <- KRAS_status %>% 
  group_by(sample_ID) %>% 
  summarise_all(funs(paste(na.omit(.), collapse = "/")))

# Clean up combined annotations
KRAS_status$KRAS_status <- gsub("AMP/|/AMP", "", KRAS_status$KRAS_status)
KRAS_status$KRAS_variant <- gsub("AMP/|/AMP", "", KRAS_status$KRAS_variant)
KRAS_status$KRAS_variant2 <- gsub("KRAS amp/|/KRAS amp", "", KRAS_status$KRAS_variant2)
KRAS_status[grep("missense", KRAS_status$KRAS_status),]$KRAS_status <- "missense"
KRAS_status[grep("Other KRAS mut", KRAS_status$KRAS_variant2),]$KRAS_variant2 <- "Other KRAS mut"

# Annotate other pathway genes
other_genes <- ALL_driver_events[ALL_driver_events$gene %in% 
                                 c("BRAF", "HRAS", "ERBB2", "EGFR", "MAP2K1", "PTEN"),]
other_genes <- unique(other_genes[,c("sample_ID", "gene", "Genomic_alterations")])
other_genes$pathway_status <- paste(other_genes$gene, other_genes$Genomic_alterations, sep = "_")
other_genes[other_genes$Genomic_alterations == "Neutral",]$pathway_status <- NA

# Remove passenger PTEN missense variants
other_genes <- other_genes[!other_genes$pathway_status == "PTEN_missense",]
other_genes$gene <- NULL
other_genes$Genomic_alterations <- NULL

other_genes <- other_genes %>% 
  group_by(sample_ID) %>% 
  summarise_all(funs(paste(na.omit(.), collapse = "/")))

other_genes <- merge(SAMPLES, other_genes, by = "sample_ID", all.x = T)
other_genes[is.na(other_genes$pathway_status),]$pathway_status <- "WT"

# Categorize pathway alterations
other_genes$pathway_status2 <- "Other"
other_genes[grep("MAP2K1_missense", other_genes$pathway_status),]$pathway_status2 <- "MAP2K1_mut"
other_genes[grep("ERBB2_Amplification", other_genes$pathway_status),]$pathway_status2 <- "ERBB2_amp"
other_genes[grep("EGFR_Amplification", other_genes$pathway_status),]$pathway_status2 <- "EGFR_amp"
other_genes[grep("PTEN_Large deletion|PTEN_SV", other_genes$pathway_status),]$pathway_status2 <- "PTEN_del"
other_genes[grep("BRAF", other_genes$pathway_status),]$pathway_status2 <- "BRAF_mut"
other_genes[grep("WT", other_genes$pathway_status),]$pathway_status2 <- "wt"

# Simplified BRAF classification
other_genes$pathway_status3 <- "Other"
other_genes[grep("BRAF", other_genes$pathway_status),]$pathway_status3 <- "BRAF_mut"
other_genes[grep("WT", other_genes$pathway_status),]$pathway_status3 <- "wt"

# Merge all annotations
CRISPR_LFC_ann <- merge(CRISPR_LFC2, KRAS_status, by = "sample_ID")
CRISPR_LFC_ann <- merge(CRISPR_LFC_ann, other_genes, by = "sample_ID")

# Filter for colorectal models
CRISPR_LFC_ann_COLO <- CRISPR_LFC_ann[CRISPR_LFC_ann$sample_ID %in% 
                                      cancer_type[cancer_type$cancer_type == "COAD/READ",]$sample_ID,]

# Set factor levels
CRISPR_LFC_ann_COLO$KRAS_variant2 <- factor(CRISPR_LFC_ann_COLO$KRAS_variant2, 
                                             levels = c("KRAS G12 mut", "Other KRAS mut","KRAS wt"))
CRISPR_LFC_ann_COLO$gene <- factor(CRISPR_LFC_ann_COLO$gene, levels = c("KRAS", "EGFR", "PTPN11"))

# Plot dependency by KRAS status (all samples)
ggplot(CRISPR_LFC_ann_COLO[CRISPR_LFC_ann_COLO$gene %in% c("EGFR", "PTPN11", "KRAS"),], 
       aes(x = KRAS_variant2, y = LFC))+
  geom_boxplot(lwd = 0.2, outlier.shape = NA) +
  geom_jitter(aes(color = pathway_status3), position=position_jitter(0.1), 
             alpha = 0.5, size = 0.5) +
  facet_grid(. ~ gene, space = "free", scales = "free") +
  scale_color_manual(values = c("red", "orangered4", "black")) +
  scale_y_continuous(limits = c(-3.5, 2.5)) +
  labs(x = NULL, y = "LFC") +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.text.x = element_text(size = 6))

# Plot dependency excluding BRAF mutants with KRAS WT
ggplot(CRISPR_LFC_ann_COLO[CRISPR_LFC_ann_COLO$gene %in% c("EGFR", "PTPN11", "KRAS") & 
                           !(CRISPR_LFC_ann_COLO$KRAS_status == "WT" & 
                             CRISPR_LFC_ann_COLO$pathway_status3 == "BRAF_mut"),], 
       aes(x = KRAS_variant2, y = LFC))+
  geom_boxplot(lwd = 0.2, outlier.shape = NA) +
  geom_jitter(position=position_jitter(0.1), alpha = 0.5, size = 0.5) +
  facet_grid(. ~ gene, space = "free", scales = "free") +
  scale_y_continuous(limits = c(-3.5, 2.5)) +
  labs(x = NULL, y = "LFC") +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        strip.text.x = element_text(size = 6))
```

## Panel b - Drug response by KRAS status
```{r}
# Load drug response data from supplementary table 9
drugs <- as.data.frame(fread('/path_to_folder/supplementary_table_9.csv'))

# Merge with KRAS status annotations
drugs <- merge(drugs, KRAS_status, by = "sample_ID")

# Set factor levels for KRAS variants
drugs$KRAS_variant2 <- factor(drugs$KRAS_variant2, 
                              levels = c("KRAS G12 mut", "Other KRAS mut", "KRAS wt"))

# Create specific G12C and G12D categories
drugs$KRAS_variant_G12C <- "Other KRAS mut"
drugs[drugs$KRAS_variant == "G12C",]$KRAS_variant_G12C <- "KRAS G12C mut"
drugs[drugs$KRAS_variant == "WT",]$KRAS_variant_G12C <- "KRAS wt"

drugs$KRAS_variant_G12D <- "Other KRAS mut"
drugs[drugs$KRAS_variant == "G12D",]$KRAS_variant_G12D <- "KRAS G12D mut"
drugs[drugs$KRAS_variant == "WT",]$KRAS_variant_G12D <- "KRAS wt"

# Statistical tests for drug responses
stat.test.afatinib <- drugs %>% 
  wilcox_test(Afatinib_uM ~ KRAS_variant2, p.adjust.method = "fdr")
stat.test.afatinib <- stat.test.afatinib[stat.test.afatinib$p.adj < 0.05,]

stat.test.gefitinib <- drugs %>% 
  wilcox_test(Gefitinib_uM ~ KRAS_variant2, p.adjust.method = "fdr")
stat.test.gefitinib <- stat.test.gefitinib[stat.test.gefitinib$p.adj < 0.05,]

# Define KRAS variant colors
mycolors_KRASvariants <- c("WT" = "grey50", "G12A" = "greenyellow", 
                           "G12C" = "springgreen3", "G12D" = "forestgreen", 
                           "G12V" = "darkseagreen1", "A146T" = "orange", 
                           "A146V" = "sienna1", "Q61H" = "purple")
```

```{r}
# Plot Afatinib response
ggplot(drugs, aes(x = KRAS_variant2, y = Afatinib_uM)) +
  geom_boxplot(lwd = 0.2, outlier.shape = NA) +
  geom_jitter(aes(color = KRAS_variant), position=position_jitter(0.1), 
             alpha = 0.5, size = 1) +
  scale_color_manual(values = mycolors_KRASvariants) +
  stat_pvalue_manual(stat.test.afatinib, label = "P = {round(p.adj, digits = 5)}", 
                     vjust = -1, y.position = 2.5, size=1, bracket.size = 0.2) +
  scale_y_log10()+
  labs(x = NULL, y = "Afatinib IC50 - log10 scale") +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.position = "none", 
        strip.text.x = element_text(size = 6))
```

```{r}
# Plot Gefitinib response
ggplot(drugs, aes(x = KRAS_variant2, y = Gefitinib_uM)) +
  geom_boxplot(lwd = 0.2, outlier.shape = NA) +
  geom_jitter(aes(color = KRAS_variant), position=position_jitter(0.1), 
             alpha = 0.5, size = 1) +
  scale_color_manual(values = mycolors_KRASvariants) +
  stat_pvalue_manual(stat.test.gefitinib, label = "P = {round(p.adj, digits = 5)}", 
                     vjust = -1, y.position = 2.5, size=1, bracket.size = 0.2) +
  scale_y_log10()+
  labs(x = NULL, y = "IC50 Gefitinib - log10 scale") +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
        legend.position = "none", 
        strip.text.x = element_text(size = 6))
```

## Panel c - Cell viability assay
```{r}
# NOTE: Data for this panel was analyzed in Prism 10
# Source file: extra information - percentages_cell_title_glo_-EGF.xlsx
# Analysis: Normalized intensities without EGF treatment
# This panel shows Western blot/cell viability data that was not analyzed in R
```

## Panel d - Reproducibility of CRISPR screens (paired organoids)
```{r}
# Extract paired organoid data (HCM-SANG-0299-C15-A vs -B)
CRISPR_LFC_pair <- CRISPR_LFC[grep("HCM-SANG-0299-C15", CRISPR_LFC$sample_ID),]
CRISPR_LFC_pair <- CRISPR_LFC_pair[CRISPR_LFC_pair$library == "v1_1",]
CRISPR_LFC_pair$library <- NULL

# Reshape to wide format for comparison
CRISPR_LFC_pair <- spread(CRISPR_LFC_pair, "sample_ID", "LFC")

# Correlation plot between paired samples
ggplot(CRISPR_LFC_pair, mapping = aes(x=`HCM-SANG-0299-C15-A`, y=`HCM-SANG-0299-C15-B`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, linetype="dashed", 
              color = "red", linewidth = 0.5) +
  stat_cor(aes(label = after_stat(r.label)), color = "red", geom = "label") +
  common_ggplot2 +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0),
        legend.position = "right")
```

## Panel e - Drug dose-response curves
```{r}
# NOTE: Data for this panel was analyzed in Prism 10
# Source file: extra information - percentages_cell_title_glo_pair.xlsx
# Model: Sigmoidal, 4PL (4-parameter logistic), X is log(concentration)
# This panel shows drug dose-response curves that were fitted in Prism
```


# Session info
```{r sessionInfo, echo=FALSE}
R version 4.5.0 (2025-04-11)
Platform: x86_64-apple-darwin20
Running under: macOS Sequoia 15.7.1

Matrix products: default
BLAS:   /System/Library/Frameworks/Accelerate.framework/Versions/A/Frameworks/vecLib.framework/Versions/A/libBLAS.dylib 
LAPACK: /Library/Frameworks/R.framework/Versions/4.5-x86_64/Resources/lib/libRlapack.dylib;  LAPACK version 3.12.1

locale:
[1] en_US.UTF-8/en_US.UTF-8/en_US.UTF-8/C/en_US.UTF-8/en_US.UTF-8

time zone: Europe/London
tzcode source: internal

attached base packages:
[1] stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
 [1] clusterProfiler_4.16.0 arrow_21.0.0           ggnewscale_0.5.2       ggpubr_0.6.1           rstatix_0.7.2         
 [6] cowplot_1.2.0          lubridate_1.9.4        forcats_1.0.0          stringr_1.5.1          dplyr_1.1.4           
[11] purrr_1.1.0            readr_2.1.5            tidyr_1.3.1            tibble_3.3.0           ggplot2_3.5.2         
[16] tidyverse_2.0.0        data.table_1.17.8     

loaded via a namespace (and not attached):
 [1] DBI_1.2.3               gson_0.1.0              rlang_1.1.6             magrittr_2.0.3          DOSE_4.2.0             
 [6] compiler_4.5.0          RSQLite_2.4.2           png_0.1-8               vctrs_0.6.5             reshape2_1.4.4         
[11] pkgconfig_2.0.3         crayon_1.5.3            fastmap_1.2.0           backports_1.5.0         XVector_0.48.0         
[16] tzdb_0.5.0              enrichplot_1.28.4       UCSC.utils_1.4.0        bit_4.6.0               xfun_0.52              
[21] cachem_1.1.0            aplot_0.2.8             GenomeInfoDb_1.44.1     jsonlite_2.0.0          blob_1.2.4             
[26] BiocParallel_1.42.1     broom_1.0.9             parallel_4.5.0          R6_2.6.1                stringi_1.8.7          
[31] RColorBrewer_1.1-3      car_3.1-3               GOSemSim_2.34.0         Rcpp_1.1.0              assertthat_0.2.1       
[36] knitr_1.50              ggtangle_0.0.7          R.utils_2.13.0          IRanges_2.42.0          igraph_2.1.4           
[41] Matrix_1.7-3            splines_4.5.0           timechange_0.3.0        tidyselect_1.2.1        qvalue_2.40.0          
[46] rstudioapi_0.17.1       dichromat_2.0-0.1       abind_1.4-8             codetools_0.2-20        lattice_0.22-6         
[51] plyr_1.8.9              treeio_1.32.0           Biobase_2.68.0          withr_3.0.2             KEGGREST_1.48.1        
[56] evaluate_1.0.4          gridGraphics_0.5-1      Biostrings_2.76.0       ggtree_3.16.3           pillar_1.11.0          
[61] carData_3.0-5           stats4_4.5.0            ggfun_0.2.0             generics_0.1.4          S4Vectors_0.46.0       
[66] hms_1.1.3               tidytree_0.4.6          scales_1.4.0            glue_1.8.0              lazyeval_0.2.2         
[71] tools_4.5.0             fgsea_1.34.2            ggsignif_0.6.4          fs_1.6.6                fastmatch_1.1-6        
[76] grid_4.5.0              ape_5.8-1               AnnotationDbi_1.70.0    colorspace_2.1-1        nlme_3.1-168           
[81] patchwork_1.3.1         GenomeInfoDbData_1.2.14 Formula_1.2-5           cli_3.6.5               gtable_0.3.6           
[86] R.methodsS3_1.8.2       yulab.utils_0.2.0       digest_0.6.37           BiocGenerics_0.54.0     ggrepel_0.9.6          
[91] ggplotify_0.1.2         farver_2.1.2            memoise_2.0.1           R.oo_1.27.1             lifecycle_1.0.4        
[96] httr_1.4.7              GO.db_3.21.0            bit64_4.6.0-1   
 
```
